Parsing Postfix log files

### 2008-03-04 doc/logparser.tex
More corrections from proofreading.

Most mail server administrators will have performed some basic processing
of the logs produced by their mail server at one time or another, whether
it was to debug a problem, explain to a user why their mail is being
rejected, or check whether new anti-spam measures are working.  The more
adventurous will have generated statistics to show how many hits each of
their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate the point to management or
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each log line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.\footnote{A
history of all changes made to Postfix is distributed with the source code,
available from \url{http://www.postfix.org/}} Further complication arises
because the set of rejection messages is not fixed: new messages can be
added by the administrator with custom checks; every \RBL{} returns a
different explanatory message; policy servers may log different messages
depending on the characteristics of the connection; there are many ways in
which the log lines may differ between servers, even within the same
organisation --- servers may be configured differently, or running
different version of Postfix.  This paper documents the difficult process
of parsing Postfix logs, presenting a program which parses logs and places
the resulting data into a database for later use.  The gathered data can
then be used to optimise current anti-spam measures, provide a baseline to
test new anti-spam measures against, or to produce statistics showing how
effective those measures are.  There are numerous other uses for such data:
improving server performance by identifying troublesome destinations and
reconfiguring appropriately; identifying regular high volume uses (e.g.\
customer newsletters) and restricting those uses to off-peak times;
detecting virus outbreaks which propagate via email; as a base for billing
customers on a shared server.  Preserving the raw data enables users to
develop a multitude of uses far beyond those conceived of by the author.

@@

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of \numberOFlogFILES{} log files taken from a mail server
handling mail for over 700 users.  Coverage is described both in terms of
the fraction of lines parsed and the fraction of mails and connections
successfully reconstructed by the parsing algorithm, including dealing with
false negatives and a discussion of the difficulties in identifying false
positives.  A random sampling of log lines was parsed, and the correctness
of the results manually verified, as part of determining the coverage of
the parser.

@@

These database queries yield summary statistics about the efficiency of
spam avoidance techniques that is far less feasible to assess directly from
log files without prior pre-processing into a database along the lines
proposed, implemented and tested herein.

@@

    \item [actions] Are invoked to deal with a log line once the line has
        been identified by the rules: modifying data structures, handing
        complications, and saving data to the database.  The actions
        perform the work of reconstructing the journey a mail takes through
        Postfix.  Full details of actions can be found in
        \sectionref{parsing-algorithm}, particularly
        \sectionref{actions-in-detail} and \sectionref{adding new actions}.

@@

Decoupling the parsing rules from the associated actions allows new rules
to be written and tested without requiring modifications to the parser
source code (significantly lowering the barrier to entry for new or casual
users who need to parse new log lines), and greatly simplifies both actions
and rules.  Decoupling also creates a clear separation of functionality:
rules handle low level details of identifying log lines and extracting data
from a line, whereas the actions handle the higher level details of
following the path a mail takes through Postfix, assembling the required
data before storing it, dealing with complications arising, etc.

@@

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Although
this may seem like a trivial point, is it substantially more difficult to
understand a program's entire parsing algorithm, identify the correct
location to change and make the appropriate changes, versus adding a new
rule with the action to invoke, a \regex{} to match the lines, and the
specification of the data to extract.  Bear in mind that the changes must
be made without adversely affecting existing parsing, particularly as there
may be edge cases which are not immediately obvious.\footnote{See
\sectionref{yet-more-aborted-delivery-attempts} for a complication which
occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
changes to the parser's code also complicates upgrades, as the changes
must be preserved during the upgrade, and may clash with changes made by
the developer.\footnote{See \sectionref{adding new actions} for how to add
new actions}.  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare case that the new actions require support from
other sections of the code will more extensive changes be required.

@@

Rules are discussed in detail in \sectionref{rules}, but the structure of
the rules table is covered here.  Rules are created by the user, not the
parser, and will not be modified by the parser (except for the hits and
hits\_total fields).  Rules parse the individual lines, extracting data to
be saved in the connections and results tables, and specifying action to
take for that line.

@@

    \item [postfix\_action] This is the action Postfix must have taken to
        generate this line.  This field is mostly ignored, but two values
        (IGNORED and INFO) have special meaning, as described below in the
        list of typical values.\label{postfix_action}

@@

        Uninteresting lines are parsed so that any lines the parser isn't
        capable of handling become immediately obvious errors.

    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
        log lines the rule applies to.  This avoids needlessly trying rules
        which won't match the line, or worse, might match unintentionally.
        Rules whose program is \texttt{*} will be tried against any lines
        which aren't parsed by program specific rules.

@@

    \item [action] The action that will be invoked when this rule matches a
        line; a full list can be found in \sectionref{actions-in-detail}.

@@

    \item [timestamp] The time the line was logged, in seconds since the
        epoch.

@@

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, followed by the problems caused by
overlapping rules, and techniques to detect and deal with such rules.  An
example rule and a log line it would match are provided, plus a description
of how the fields in the rule\footnote{The details of the table containing
the rules have already been described in \sectionref{rule attributes}.} are
used when matching a line and subsequently performing the requested action.
The section continues with a discussion of rule efficiency concerns,
referring to the graphs in \sectionref{graphs}, and finishes with an
explanation of the algorithm used to generate new \regexes{} from unparsed
lines.

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.\footnote{There are also generic rules which are used when
        parsing log lines produced by any Postfix program, but only if
        there are also rules specific to that program, and those rules have
        already have been tried and failed on the current line.}  Any given
        rule will only be used to parse a subset of the log lines, and any
        given log line will only be parsed by a subset of the rules.

    \item The first matching rule wins: no further rules are tried against
        that line, but there is a facility for ordering the rules so that
        more specific rules can be tried first.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's \regex{} and determine if any overlap.  The
author has not attempted to do this: such a project by itself would
probably qualify for a PhD, and may involve solving the Halting
Problem~\cite{Wikipedia-halting-problem} and circumventing the
Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line may change between log
files, and the first matching rule wins.  Overlapping rules are frequently
a requirement, allowing a more specific rule to match some lines and a more
general rule to match the majority, e.g.\ separating \SMTP{} delivery to
specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches when parsing the previous log file.  Negative
priorities may be useful for catchall rules.

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the line a match using the \regex{} will be
        attempted against the line; if the match is successful the action
        will be executed, if not the next rule will be tried.  If the
        program-specific rules don't match the log line, the parser will
        fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.

@@

    \item [queueid:] The index of the capture in the \regex{} which
        supplies the queueid, or zero if the line does not contain a
        queueid.  This allows the correct mail can be found by queueid and
        actions performed on it.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (generated daily on
a mail server handling mail for approximately 700 users --- large scale
mail servers would have much larger log files on a daily basis).  When
generating the data for the graphs included in \sectionref{graphs},
\numberOFlogFILES{} log files (totaling 10.08 GB, 60.72 million lines) were
each parsed 10 times, the first run discarded, and the execution time for
the remaining 9 runs averaged.  The first run is discarded for two reasons:

@@

An important property of a parser is how execution time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
Graph~\refwithpage{execution time vs file size vs number of lines graph}
shows the execution time in seconds, file size in MB and tens of thousands
of lines per log file.  All three lines run roughly in parallel, giving a
visual impression that the algorithm scales linearly with input size.  This
impression is borne out by graph~\refwithpage{execution time vs file size
vs number lines factor}, which plots the ratio of file size vs execution
time and ratio of number of lines vs execution time (higher is
better).\footnote{Table~\refwithpage{execution time vs file size vs number
lines factor table} shows the ratios for different breakdowns of the log
files.}  As the reader can see the ratios are quite tightly banded, showing
that the algorithm scales linearly: the much larger log files between
points 60 and 70 on the X axis in graph~\refwithpage{execution time vs file
size vs number of lines graph} actually cause the ratio to increase, rather
than decrease.  The strange behaviour where larger log files are parsed
more efficiently is explained fully in \sectionref{Why are there dips in
the graphs?}.

@@

The logs produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \RBL{}
services, policy servers, or custom rejection messages.  To facilitate easy
parsing of new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions is difficult, but adding new rules to
parse new rejection messages is trivial (and also occurs much more
frequently).  The implementation provides a program to ease the process of
creating new rules from unparsed lines, based on the algorithm developed by
Risto Vaarandi~\cite{risto-vaarandi} for his \SLCT{}~\cite{slct-paper}.
The differences between the two algorithms will be outlined as part of the
general explanation below.

The core of the algorithm is quite simple: log lines are generally created
by substituting variable words into a fixed pattern, and analysis of the
frequency with which each word occurs can be used to determine whether the
word is variable or part of the fixed pattern.  This classification can be
used to group similar log lines and generate a \regex{} to match each group
of lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained while writing rules and performs a large number of
        substitutions on the input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        which the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this phase is to utilise existing knowledge to create more accurate
        \regexes{}.  The new log lines are written to a temporary file,
        which all subsequent stages use instead of the original input file.

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash from a small range of values for each word
        of each line, incrementing a counter for each hash.  The counters
        will be used in the next pass to filter out words: if the word's
        hash does not have a high frequency, the word itself cannot have a
        high frequency, and there is no need to maintain a counter for it,
        reducing the number of counters and thus the program's memory
        consumption.

    \item [Calculate word frequencies]  The position of words within a line
        is important: a common word does not indicate similarity between
        lines unless it occupies the same position within both
        lines.\footnote{If a variable term within a line contains spaces,
        it will appear to the algorithm as two words rather than one.  This
        will alter the position of subsequent words, so a word occurring in
        different positions in two log lines \textit{may\/} indicate
        similarity, but the algorithm does not attempt to deal with this
        possibility.}  The algorithm maintains a counter for each
        \textit{(word, word's position within the line)\/} tuple,
        incrementing it each time that word occurs in that position.

@@

    \item [Classify words based on their frequency]  The frequency of each
        \textit{(word, word's position within the line)\/} tuple is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all lines is generally a good starting point) it
        is classified as a fixed word, otherwise it is classified as a
        variable term.  If a variable term appears sufficiently often it
        will be classified as a fixed term, but that should be noticed by
        the user when reviewing the new \regexes{}.  Variable terms are
        replaced by \texttt{.+}, which means to match zero or more of any
        character.

    \item [Build regexes]  The words are reassembled to produce a \regex{}
        matching the line, and a counter is maintained for each \regex{}.
        Contiguous sequences of \texttt{.+} in the newly reassembled
        \regexes{} are collapsed to a single \texttt{.+}; any resulting
        duplicate \regexes{} are combined, and their counters added.  If
        the frequency of a \regex{} is lower than the threshold supplied by
        the user the \regex{} is discarded.\footnote{This is a second
        threshold, independent of the threshold used to differentiate
        between fixed and variable words, but once again 1\% of input lines
        is a good starting point; obviously the threshold depends on the
        number and type of input lines.}  The new \regexes{} are printed
        for the user to add to the database, either as new rules or merged
        into the \regexes{} of existing rules; the counter for each
        \regex{} is also printed, giving the user an indication of how many
        of the input lines that \regex{} should match.  Discarding
        \regexes{} will result in some of the input lines not being
        matched; this utility should be run again once the unmatched lines
        have been isolated by running the parser, including the new
        \regexes{}, with the same input which produced the original set of
        unparsed lines.

@@

A second utility is also provided which reads a list of new \regexes{} and
the input given to the first utility.  It tries to match each input line
against each \regex{}, counting the number of lines which match each
\regex{}, warning the user if an input line is matched by more than one
\regex{}, and additionally warning if an input line is not matched by any
\regex{}.  It displays a summary of how many input lines each \regex{}
matched, comparing it to the expected number of matches; this provides the
user with an easy method of checking if the \regexes{} produced by the
first utility are correctly matching the input lines they are based upon.
A future version of this utility will also group input lines by \regex{},
so the user can tweak the \regexes{} if required.

@@

    \item This section finished with an explanation of the algorithm used
        to generate a new \regex{} from unparsed lines.

@@

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation line!  Thus mails flagged as
coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as ready for
entry and subsequently entered by the parent mail once it has been
identified.

@@

    \item Check if there are any rules for the program which produced the
        log line; if not then skip the line.

    \item Try each rule until a matching rule is found.

@@

    \item Each program of interest must have at least one rule or its log
        lines will be silently skipped; in the extreme case of zero rules
        the parser would happily skip every log line.  There may be any
        number of log lines from other programs intermingled in the log
        file, and there are some Postfix programs which don't produce any
        log lines of interest.

    \item There must be a rule to match each different line produced by
        each program; if a line is not successfully matched the parser will
        issue a warning.  Rules should be as specific and tightly bound as
        possible to ensure accurate parsing:\footnote{A rule which matches
        zero or more of any character will successfully parse every log
        line, but not in a meaningful way.} most log lines contain fixed
        strings and have a rigid pattern, so this is not a problem.

@@

Full coverage of log lines is easy to achieve yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files (at the time
of writing the parser has \numberOFrules{} rules, fully parsing
\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
rules are easy to add.  Maintaining full coverage is hard because other
servers have different restrictions with custom messages, \RBL{} messages
change over time, major releases of Postfix change warning messages
(usually adding more information), etc.,\ so over time the log lines drift
and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
for each rule over all \numberOFlogFILES{} log files; it's obvious that a
small number of rules match the vast majority of the lines, and more than
half the rules match fewer than 100 times.

Warnings are issued for any lines which are not parsed; no warnings are
issued for unparsed lines while testing with the \numberOFlogFILES{} test
log files, so it can be safely concluded that there are zero false
negatives.  False positives are harder to quantify: short of examining each
of the 60,721,709 log lines and determining which \regex{} parsed it, there
is no way to be sure that every line was parsed by the correct \regex{},
making it impossible to quantify the false positive rate; however a random
sample of 6039 log lines was parsed and the results checked manually to
ensure that the correct \regex{} parsed each line.  The sample was
generated by running the following command:

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (many of the
complications discussed in \sectionref{additional complications} were
detected in this way).  The parser issues warnings when it detects any
errors, some of which may alert the user to a problem, e.g.\ when a queueid
is reused before the previous mail is fully dealt with, when a queueid or
\pid{} is not found,\footnote{There will often be warnings about a missing
queueid or \pid{} in the first few hundred or thousand log lines because
the earlier log lines for those connections or mails are in the previous
log file; loading the saved state from the previous log file will solve
this problem.} or when there are problems tracking a child mail (see
\sectionref{tracking re-injected mail}).  There should be few or no
warnings when parsing, and when finished parsing the state table should
only contain entries for mails which had yet to be delivered when the log
files ended, or started before the log files began.

@@

    \item 1 connection which started only seconds before the logs ended and
        had not yet completed.

@@

    \item Queueids and \pids{} naturally group together log lines belonging
        to one mail or connection, respectively; it is extremely unlikely
        that a log line would be associated with the wrong connection.

>>>>

Most mail server administrators will have performed some basic processing
of the logs produced by their mail server at one time or another, whether
it was to debug a problem, explain to a user why their mail is being
rejected, or check whether new anti-spam measures are working.  The more
adventurous will have generated statistics to show how many hits each of
their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate the point to management or
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this type of processing is the
unstructured nature of Postfix logs, where each log line was added on an ad
hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every \RBL{}
returns a different explanatory message; policy servers may log different
messages depending on the characteristics of the connection; there are many
ways in which the log lines may differ between servers, even within the
same organisation --- servers may be configured differently, or running
different version of Postfix.  This paper documents the difficult process
of parsing Postfix logs, presenting a program which parses logs and places
the resulting data into a database for later use.  The gathered data can
then be used to optimise current anti-spam measures, provide a baseline to
test new anti-spam measures against, or to produce statistics showing how
effective those measures are.  There are numerous other uses for such data:
improving server performance by identifying troublesome destinations and
reconfiguring appropriately; identifying regular high volume uses (e.g.\
customer newsletters) and restricting those uses to off-peak times;
detecting virus outbreaks which propagate via email; as a base for billing
customers on a shared server.  Preserving the raw data enables users to
develop a multitude of uses far beyond those conceived of by the author.

@@

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of \numberOFlogFILES{} log files taken from a mail server
handling mail for over 700 users.  Coverage is described both in terms of
the fraction of log lines parsed and the fraction of mails and connections
successfully reconstructed by the parsing algorithm, including dealing with
false negatives and a discussion of the difficulties in identifying false
positives.  A random sampling of log lines was parsed, and the correctness
of the results manually verified, as part of determining the coverage of
the parser.

@@

These database queries yield summary statistics about the efficiency of
spam avoidance techniques that is far less feasible to assess directly from
log files without prior pre-processing into a database in the fashion
proposed, implemented and tested herein.

@@

    \item [actions] Are invoked to deal with a log line once it has been
        identified by the rules: modifying data structures, handing
        complications, and saving data to the database.  The actions
        perform the work of reconstructing the journey a mail takes through
        Postfix.  Full details of actions can be found in
        \sectionref{parsing-algorithm}, particularly
        \sectionref{actions-in-detail} and \sectionref{adding new actions}.

@@

Decoupling the parsing rules from the associated actions allows new rules
to be written and tested without requiring modifications to the parser
source code (significantly lowering the barrier to entry for new or casual
users who need to parse new log lines), and greatly simplifies both actions
and rules.  Decoupling also creates a clear separation of functionality:
rules handle low level details of identifying log lines and extracting data
from a log line, whereas the actions handle the higher level details of
following the path a mail takes through Postfix, assembling the required
data before storing it, dealing with complications arising, etc.

@@

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Although
this may seem like a trivial point, is it substantially more difficult to
understand a program's entire parsing algorithm, identify the correct
location to change and make the appropriate changes, versus adding a new
rule with the action to invoke, a \regex{} to match the log lines, and the
specification of the data to extract.  Bear in mind that the changes must
be made without adversely affecting existing parsing, particularly as there
may be edge cases which are not immediately obvious.\footnote{See
\sectionref{yet-more-aborted-delivery-attempts} for a complication which
occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
changes to the parser's code also complicates upgrades, as the changes
must be preserved during the upgrade, and may clash with changes made by
the developer.\footnote{See \sectionref{adding new actions} for how to add
new actions}.  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare case that the new actions require support from
other sections of the code will more extensive changes be required.

@@

Rules are discussed in detail in \sectionref{rules}, but the structure of
the rules table is covered here.  Rules are created by the user, not the
parser, and will not be modified by the parser (except for the hits and
hits\_total fields).  Rules parse the individual log lines, extracting data
to be saved in the connections and results tables, and specifying action to
take for that log line.

@@

    \item [postfix\_action] This is the action Postfix must have taken to
        generate this log line.  This field is mostly ignored, but two
        values (IGNORED and INFO) have special meaning, as described below
        in the list of typical values.\label{postfix_action}

@@

        Uninteresting log lines are parsed so that any lines the parser
        isn't capable of handling become immediately obvious errors.

    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
        log lines the rule applies to.  This avoids needlessly trying rules
        which won't match the log line, or worse, might match
        unintentionally.  Rules whose program is \texttt{*} will be tried
        against any log lines which aren't parsed by program specific
        rules.

@@

    \item [action] The action that will be invoked when this rule matches a
        log line; a full list can be found in
        \sectionref{actions-in-detail}.

@@

    \item [timestamp] The time the log line was logged, in seconds since
        the epoch.

@@

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, followed by the problems caused by
overlapping rules, and techniques to detect and deal with such rules.  An
example rule and a log line it would match are provided, plus a description
of how the fields in the rule\footnote{The details of the table containing
the rules have already been described in \sectionref{rule attributes}.} are
used when matching a log line and subsequently performing the requested
action.  The section continues with a discussion of rule efficiency
concerns, referring to the graphs in \sectionref{graphs}, and finishes with
an explanation of the algorithm used to generate new \regexes{} from
unparsed lines.

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.\footnote{There are also generic rules which are used when
        parsing log lines produced by any Postfix program, but only if
        there are also rules specific to that program, and those rules have
        already have been tried and failed on the current log line.}  Any
        given rule will only be used to parse a subset of the log lines,
        and any given log line will only be parsed by a subset of the
        rules.

    \item The first matching rule wins: no further rules are tried against
        that log line, but there is a facility for ordering the rules so
        that more specific rules can be tried first.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's \regex{} and determine if any overlap.  The
author has not attempted to do this: such a project by itself would
probably qualify for a PhD, and may involve solving the Halting
Problem~\cite{Wikipedia-halting-problem} and circumventing the
Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line may change between log
files, and the first matching rule wins.  Overlapping rules are frequently
a requirement, allowing a more specific rule to match some log lines and a
more general rule to match the majority, e.g.\ separating \SMTP{} delivery
to specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches when parsing the previous log file.  Negative
priorities may be useful for catchall rules.

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the log line a match using the \regex{} will
        be attempted against the log line; if the match is successful the
        action will be executed, if not the next rule will be tried.  If
        the program-specific rules don't match the log line, the parser
        will fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.

@@

    \item [queueid:] The index of the capture in the \regex{} which
        supplies the queueid, or zero if the log line does not contain a
        queueid.  This allows the correct mail can be found by queueid and
        actions performed on it.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (generated daily on
a mail server handling mail for approximately 700 users --- large scale
mail servers would have much larger log files on a daily basis).  When
generating the data for the graphs included in \sectionref{graphs},
\numberOFlogFILES{} log files (totaling 10.08 GB, 60.72 million log lines)
were each parsed 10 times, the first run discarded, and the execution time
for the remaining 9 runs averaged.  The first run is discarded for two
reasons:

@@

An important property of a parser is how execution time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
Graph~\refwithpage{execution time vs file size vs number of lines graph}
shows the execution time in seconds, file size in MB and tens of thousands
of log lines per log file.  All three lines run roughly in parallel, giving
a visual impression that the algorithm scales linearly with input size.
This impression is borne out by graph~\refwithpage{execution time vs file
size vs number lines factor}, which plots the ratio of file size vs
execution time and ratio of number of log lines vs execution time (higher
is better).\footnote{Table~\refwithpage{execution time vs file size vs
number lines factor table} shows the ratios for different breakdowns of the
log files.}  As the reader can see the ratios are quite tightly banded,
showing that the algorithm scales linearly: the much larger log files
between points 60 and 70 on the X axis in graph~\refwithpage{execution time
vs file size vs number of lines graph} actually cause the ratio to
increase, rather than decrease.  The strange behaviour where larger log
files are parsed more efficiently is explained fully in \sectionref{Why are
there dips in the graphs?}.

@@

The logs produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \RBL{}
services, policy servers, or custom rejection messages.  To facilitate easy
parsing of new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions is difficult, but adding new rules to
parse new rejection messages is trivial (and also occurs much more
frequently).  The implementation provides a program to ease the process of
creating new rules from unparsed log lines, based on the algorithm
developed by Risto Vaarandi~\cite{risto-vaarandi} for his
\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
be outlined as part of the general explanation below.

The core of the algorithm is quite simple: log lines are generally created
by substituting variable words into a fixed pattern, and analysis of the
frequency with which each word occurs can be used to determine whether the
word is variable or part of the fixed pattern.  This classification can be
used to group similar log lines and generate a \regex{} to match each group
of log lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained while writing rules and performs a large number of
        substitutions on the log input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        which the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this phase is to utilise existing knowledge to create more accurate
        \regexes{}.  The new log lines are written to a temporary file,
        which all subsequent stages use instead of the original input file.

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash from a small range of values for each word
        of each log line, incrementing a counter for each hash.  The
        counters will be used in the next pass to filter out words: if the
        word's hash does not have a high frequency, the word itself cannot
        have a high frequency, and there is no need to maintain a counter
        for it, reducing the number of counters and thus the program's
        memory consumption.

    \item [Calculate word frequencies]  The position of words within a log
        line is important: a common word does not indicate similarity
        between log lines unless it occupies the same position within both
        log lines.\footnote{If a variable term within a line contains
        spaces, it will appear to the algorithm as two words rather than
        one.  This will alter the position of subsequent words, so a word
        occurring in different positions in two log lines \textit{may\/}
        indicate similarity, but the algorithm does not attempt to deal
        with this possibility.}  The algorithm maintains a counter for each
        \textit{(word, word's position within the log line)\/} tuple,
        incrementing it each time that word occurs in that position.

@@

    \item [Classify words based on their frequency]  The frequency of each
        \textit{(word, word's position within the log line)\/} tuple is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all log lines is generally a good starting point)
        it is classified as a fixed word, otherwise it is classified as a
        variable term.  If a variable term appears sufficiently often it
        will be classified as a fixed term, but that should be noticed by
        the user when reviewing the new \regexes{}.  Variable terms are
        replaced by \texttt{.+}, which means to match zero or more of any
        character.

    \item [Build regexes]  The words are reassembled to produce a \regex{}
        matching the log line, and a counter is maintained for each
        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
        resulting duplicate \regexes{} are combined, and their counters
        added.  If the frequency of a \regex{} is lower than the threshold
        supplied by the user the \regex{} is discarded.\footnote{This is a
        second threshold, independent of the threshold used to
        differentiate between fixed and variable words, but once again 1\%
        of input log lines is a good starting point; obviously the
        threshold depends on the number and type of input log lines.}  The
        new \regexes{} are printed for the user to add to the database,
        either as new rules or merged into the \regexes{} of existing
        rules; the counter for each \regex{} is also printed, giving the
        user an indication of how many of the input log lines that \regex{}
        should match.  Discarding \regexes{} will result in some of the
        input log lines not being matched; this utility should be run again
        once the unmatched log lines have been isolated by running the
        parser, including the new \regexes{}, with the same input which
        produced the original set of unparsed log lines.

@@

A second utility is also provided which reads a list of new \regexes{} and
the input given to the first utility.  It tries to match each input log
line against each \regex{}, counting the number of log lines which match
each \regex{}, warning the user if an input log line is matched by more
than one \regex{}, and additionally warning if an input log line is not
matched by any \regex{}.  It displays a summary of how many input log lines
each \regex{} matched, comparing it to the expected number of matches; this
provides the user with an easy method of checking if the \regexes{}
produced by the first utility are correctly matching the input log lines
they are based upon.  A future version of this utility will also group
input log lines by \regex{}, so the user can tweak the \regexes{} if
required.

@@

    \item This section finished with an explanation of the algorithm used
        to generate a new \regex{} from unparsed log lines.

@@

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation log line!  Thus mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as ready for
entry and subsequently entered by the parent mail once it has been
identified.

@@

    \item Check if there are any rules for the Postfix component which
        produced the log line; if not then skip the log line.

    \item Try each rule until a matching rule is found; if no match is
        issue a warning and move to the next log line.

@@

    \item Each Postfix component whose log lines are of interest must have
        at least one rule or its log lines will be silently skipped; in the
        extreme case of zero rules the parser would happily skip every log
        line.  There may be any number of log lines from other programs
        intermingled in the log file, and there are some Postfix programs
        which don't produce any log lines of interest.

    \item There must be a rule to match each different log line produced by
        each program; if a log line is not successfully matched the parser
        will issue a warning.  Rules should be as specific and tightly
        bound as possible to ensure accurate parsing:\footnote{A rule which
        matches zero or more of any character will successfully parse every
        log line, but not in a meaningful way.} most log lines contain
        fixed strings and have a rigid pattern, so this is not a problem.

@@

Full coverage of log lines is easy to achieve yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files (at the time
of writing the parser is tested with \numberOFrules{} rules, fully parsing
\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
rules are easy to add.  Maintaining full coverage is hard because other
servers have different restrictions with custom messages, \RBL{} messages
change over time, major releases of Postfix change warning messages
(usually adding more information), etc.,\ so over time the log lines drift
and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
for each rule over all \numberOFlogFILES{} log files; it's obvious that a
small number of rules match the vast majority of the lines, and more than
half the rules match fewer than 100 times.

Warnings are issued for any log lines which are not parsed; no warnings are
issued for unparsed log lines while testing with the \numberOFlogFILES{}
test log files, so it can be safely concluded that there are zero false
negatives.  False positives are harder to quantify: short of examining each
of the 60,721,709 log lines and determining which \regex{} parsed it, there
is no way to be sure that every line was parsed by the correct \regex{},
making it impossible to quantify the false positive rate; however a random
sample of 6039 log lines was parsed and the results checked manually to
ensure that the correct \regex{} parsed each log line.  The sample was
generated by running the following command:

@@

Coverage of mails is much more difficult to determine accurately than
coverage of log lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (many of the
complications discussed in \sectionref{additional complications} were
detected in this way).  The parser issues warnings when it detects any
errors, some of which may alert the user to a problem, e.g.\ when a queueid
is reused before the previous mail is fully dealt with, when a queueid or
\pid{} is not found in the state tables,\footnote{There will often be
warnings about a missing queueid or \pid{} in the first few hundred or
thousand log lines because the earlier log lines for those connections or
mails are in the previous log file; loading the saved state from the
previous log file will solve this problem.} or when there are problems
tracking a child mail (see \sectionref{tracking re-injected mail}).  There
should be few or no warnings when parsing, and when finished parsing the
state table should only contain entries for mails which had yet to be
delivered when the log files ended, or were accepted before the log files
began.

@@

    \item 1 connection which started only seconds before the logs ended and
        had not yet been fully transferred from client to server.

@@

    \item Queueids and \pids{} naturally group together log lines belonging
        to one mail or connection respectively; it is extremely unlikely
        that a log line would be associated with the wrong connection.

    \item When dealing with the complications described in
        \sectionref{complications} and \sectionref{additional
        complications} the solutions are as specific and restrictive as
        possible, with the goal of minimising the number of false
        positives.  In addition the solution to the \textit{Out of order
        log files\/} complication described in \sectionref{out of order log
        lines} imposes conditions which each reassembled mail must comply
        with to be acceptable.
